A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.Rmd
Finished editing Analysis, improved the presentation of some figures, added alternate models for reference, for sensitivity analysis. Next step: Results and discussion, and conclusions.

---
title: A spatial analysis of the environmental correlates of COVID-19 incidence in Spain
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Fernando A. Lopez
    email: fernando.lopez@upct.es
    affiliation: Universidad Politecnica de Cartagena
  - name: Tatiane Menezes
    email: tatiane.menezes@ufpe.br
    affiliation: Departamento de Economia
  - name: Renata Cavalcanti
    email: renata.vcsantos@gmail.com
    affiliation: Nucleo de Pesquisa
  - name: Maira Galdino da Rocha Pitta
    email: mgrpitta@ufpe.br
    affiliation: Nucleo de Pesquisa
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, 1281 Main St W, Hamilton, ON, L8S 4K1, Canada
  - code: Universidad Politecnica de Cartagena
    address: Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3, 30201, Cartagena, Murcia, Spain
  - code: Departamento de Economia
    address: Departamento de Economia, Universidade Federal de Pernambuco, Av dos Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil  
  - code: Nucleo de Pesquisa
    address: Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av. Prof. Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP 50670-901, Brazil  
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the globe, causing enormous stress to health systems and prompting social distance guidelines and mandates to arrest its progress. While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees. How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature, humidity, and hours of sunshine. In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach allows us to model the incidence of reported cases of the disease per 100,000 population, as a function of temperature and humidity, while controlling for GDP per capita, population density, percentage of older adults in the population, and presence of mass transit systems. An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process. Our results indicate that incidence of the disease is lower at higher temperatures and higher levels of humidity, although coefficients for this variable are significant only in some equations. Sunshine, in contrast, displays a positive association with incidence of the disease. Our control variables also yield interesting insights. Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.    
journal: "Geographical Analysis"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

@@

Context, Data, and Methods
============

@@

The first variable that we consider is GDP per capita. Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}. The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when its cause may still be unclear [@Zhou2016accelerated]. In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] remarks how Toronto's status as a global city was a vulnerability. In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and their potential to be further ahead in the trajectory of the pandemic. Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional]. Therefore, it is possible that wealthier regions remained relatively more active even during the lockdown. On the other hand, it is also possible the less wealthy regions have a higher proportion of workers in manual occupations that cannot telework.

The percentaje of older adults (over 65) is the second variable that we consider as a control. Early evidence suggest that the case rate mortality is higher at older ages [e.g. @Novel2020epidemiological]. It is not clear that presence of older adults necessarily translates into higher transmission in the population. The public health tool of choice in containing the spread of the disease has been social distancing. In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility]. In other words, older adults are, by preference or necessity, already in a form of social isolation. Social distancing during the pandemic may actually reinforce that condition for older adults, as the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age] suggests. Accordingly, our expectation is that in the provinces with higher proportions of older adults will tend to have similarl levels of lower social contact, particularly since age-structured social contact matrix in Spain is similar to Italy [@Prem2017projecting].

@@

The last control variable is the presence of mass transit systems in a province. Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility, namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches [@Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing. After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.

@@

## Methods: Spatial SUR

@@

\noindent where ![$\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BA%7D%20%3DI_%7BTN%7D-%5Cbf%7B%5CGamma%7D%20%5Cotimes%20W%24) is the spatially lagged dependent variable, and ![$\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7B%5CGamma%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).
-->

<!--
This specification assumes that outcome (![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24)) at location ![$s$](https://render.githubusercontent.com/render/math?math=%24s%24) and time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average (![$Wy_{st}$](https://render.githubusercontent.com/render/math?math=%24Wy_%7Bst%7D%24)) of the outcome in neighboring provinces, with neighborhood defined by matrix ![$W$](https://render.githubusercontent.com/render/math?math=%24W%24) of spatial weights. In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way. The coefficients of the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identify the intensity and the sign of the contagion effect. It is possible test the null hypothesis of identical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)). The correspond Wald test is available in the `R` package `spsur`. 
-->

The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019). Another alternative methodologies could be use. By example, a dynamic spatial panel methodology with fixed spatial an temporal effects (e.g. Elhorst 2014, Cap. 4), but those models don't take account correlation between errors. Therefore, a spatial SUR approach is more reasonable for our purpose.

Analysis and Results
===================

@@

We begin with the exploratory analysis of the data.

Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of the disease, as well as its temporal progression in weekly averages. Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19. It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain. While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces. Vitoria, its main city, is the capital of the Basque country and has been the focus of efforts to develop a "Global Basque City" [@Meijers2008strategic], along with San Sebastian and Bilbao. The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava. The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.

@@

week11.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 11) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 13 - March 15") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 8, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week12.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 12) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds",
                       direction = 1) +
  ggtitle("March 16 - March 22") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 8, y = madrid_xy[2] + 0.6) +
  geom_segment(x = madrid_xy[1] - 6, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])

week13.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 13) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 23 - March 29") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid",
               x = madrid_xy[1] - 8, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week14.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 14) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 30 - April 5") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 8, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week15.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 15) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence)) +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("April 6 - April 11") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 8, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

grid.arrange(week11.plot, week12.plot, week13.plot, week14.plot, week15.plot, nrow = 3)
```

To visualize the distribution of temperature and humidity we aggregate the provinces by Autonomous Community. In Figure \ref{fig:descriptives-temperature} the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Canarias the southernmost. There is a relatively wide range of values both within and between provinces over the 30 day period examined. The top panel of the figure shows the distribution of mean temperature. The lowest mean temperature for a community on any given day is 2.8C, and the highest is 22.4C for a range of approximately 20 degrees. Likewise, there is a great deal of variability in humidity, as seen in the bottom panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6. The actual values for the provinces display somewhat more variability even.

@@

# Bind tables
env.summary <- rbind(temp.summary, humidity.summary) %>%
  mutate(Variable = factor(Variable, 
                           levels = c("Mean Temperature in Community",
                                      "Mean Humidity in Community")))

# Plot
ggplot(data = env.summary, 
       aes(x = CCAA,
           y = Value)) +
  geom_boxplot() +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Autonomous Community (sorted from north to south)") +
  ylab("Value") +
  facet_wrap(~ Variable, nrow = 2, 
             scales = "free_y")
```

```{r map-controls, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:map-controls}Spatial distribution of control variables by province", warning=FALSE, cache = TRUE}
# GDP per capita:
q <- quantile(provinces_spain$GDPpc)
gdppc.plot <- provinces_spain %>%
  mutate(GDPpc = GDPpc,
         GDPpc = case_when(GDPpc < q[2] ~ "1st",
                             GDPpc > q[2] & GDPpc <= q[3]  ~ "2nd",
                             GDPpc > q[3] & GDPpc <= q[4] ~ "3rd",
                             GDPpc > q[4] ~ "4th")) %>%
  #filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = GDPpc)) +
  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom") +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 8, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

>>>>

---
title: A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Fernando A. Lopez
    email: fernando.lopez@upct.es
    affiliation: Universidad Politecnica de Cartagena
  - name: Tatiane Menezes
    email: tatiane.menezes@ufpe.br
    affiliation: Departamento de Economia
  - name: Renata Cavalcanti
    email: renata.vcsantos@gmail.com
    affiliation: Nucleo de Pesquisa
  - name: Maira Galdino da Rocha Pitta
    email: mgrpitta@ufpe.br
    affiliation: Nucleo de Pesquisa
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, 1281 Main St W, Hamilton, ON, L8S 4K1, Canada
  - code: Universidad Politecnica de Cartagena
    address: Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3, 30201, Cartagena, Murcia, Spain
  - code: Departamento de Economia
    address: Departamento de Economia, Universidade Federal de Pernambuco, Av dos Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil  
  - code: Nucleo de Pesquisa
    address: Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av. Prof. Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP 50670-901, Brazil  
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the globe, causing enormous stress to health systems and prompting social distance guidelines and mandates to arrest its progress. While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees. How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature, humidity, and hours of sunshine. In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach allows us to model the incidence of reported cases of the disease per 100,000 population, as a function of temperature and humidity, while controlling for GDP per capita, population density, percentage of older adults in the population, and presence of mass transit systems. An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process. Our results indicate that incidence of the disease is lower at higher temperatures and higher levels of humidity, although coefficients for this variable are significant only in some equations. Sunshine, in contrast, displays a positive association with incidence of the disease. Our control variables also yield interesting insights. Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.    
journal: "Geographical Analysis"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
#numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

@@

Context and Data
============

@@

The first variable that we consider is GDP per capita. Much has been said about globalization and the spread of infectious disease\footnote{As the Globe and Mail, Canada's paper of record, put it in relation to the SARS outbreak in 2003: "Globalization means that if someone in China sneezes, someone in Toronto may one day catch a cold" (Editorial, March 28, 2003, p. A18)}. The growth in global connections has presented a challenge to spatial approaches in the initial stages of disease management, when the cause of a disease may still be unclear, but the plane has already departed [@Zhou2016accelerated]. In reference to the earlier outbreak of SARS, van Wagner [@vanWagner2008toward] chronicles how Toronto's status as a global city turned out to be a vulnerability in this respect. In our case, we think of GDP per capita as a marker of a region's relative position in a network of global cities, and its potential to be further ahead in the trajectory of the pandemic. Furthermore, wealthier regions also tend to concentrate more activities that produce non-traded goods, including building and construction [@Hallet2002regional]. Therefore, it is possible that wealthier regions remain relatively more active even during a lockdown. On the other hand, it is also possible that less wealthy regions have a higher proportion of workers in manual occupations that cannot telework, and therefore have more difficulties complying with shelter-in-place orders.

The percentaje of older adults (over 65) is the second variable that we consider as a control. Early evidence regarding COVID-19 suggests that the case rate mortality is higher at older ages [e.g. @Novel2020epidemiological]. However, it is not clear that a relatively large population of older adults necessarily translates into higher transmission of the infection. The tool of choice in containing the spread of the disease has been social distancing. In this respect, the evidence from the field of transportation is that older adults tend to travel less frequently, for shorter distances, and have higher rates of immobility [e.g., @Roorda2010trip; @Morency2011distance; @Sikder2012immobility]. In other words, many older adults are, whether by preference or otherwise, already in a form of social isolation. Social distancing during the pandemic may actually reinforce that condition for them, as suggested by the analysis of age-structured social contact in India, China, and Italy of Sing and Adhikari [-@Singh2020age]. Accordingly, our expectation is that provinces with higher percentages of older adults will tend to have similarly low levels of social contact, particularly since the age-structured matrix of social contact in Spain is similar to Italy [@Prem2017projecting].

@@

The last control variable is the presence of mass transit systems in a province. Every province in Spain offers some form of public transportation, however only five provinces have higher order systems of mass mobility (e.g. metro or subway), namely Barcelona, Madrid, Sevilla, Valencia, and Bizkaia. Public transportation has been hypothesized to relate to the spread of contagious disease by some researchers using agent-based approaches and simulation [e.g., @Perez2009agent; @Wang2010gis], and while we find scant evidence of a link in the literature, the idea is intuitively appealing. After all, unlike the isolation that a car offers to travellers, most mass transit system are cauldrons of social contact.

@@

Methods: the Spatial SUR Model
===================

@@

\noindent where ![$\bf{A} =I_{TN}-\bf{\Gamma} \otimes W$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7BA%7D%20%3DI_%7BTN%7D-%5Cbf%7B%5CGamma%7D%20%5Cotimes%20W%24) is the spatially lagged dependent variable, and ![$\bf{\Gamma} = diag(\rho_1, \cdots, \rho_T)$](https://render.githubusercontent.com/render/math?math=%24%5Cbf%7B%5CGamma%7D%20%3D%20diag(%5Crho_1%2C%20%5Ccdots%2C%20%5Crho_T)%24).

This specification assumes that outcome (![$y_{st}$](https://render.githubusercontent.com/render/math?math=%24y_%7Bst%7D%24)) at location ![$s$](https://render.githubusercontent.com/render/math?math=%24s%24) and time ![$t$](https://render.githubusercontent.com/render/math?math=%24t%24) is partially determined by the weighted average (![$Wy_{st}$](https://render.githubusercontent.com/render/math?math=%24Wy_%7Bst%7D%24)) of the outcome in neighboring provinces, with neighborhood defined by matrix ![$W$](https://render.githubusercontent.com/render/math?math=%24W%24) of spatial weights. In other words, the spatially lagged dependent variable represents a process of contagion, where the disease in neighboring provinces can spillover in a spatial way. The coefficients of the spatially lagged variable are estimated for each time period ![$\rho_t$](https://render.githubusercontent.com/render/math?math=%24%5Crho_t%24) and identify the intensity and the sign of the contagion effect. It is possible test the null hypothesis of identical levels of spatial dependence (![$\rho_i=\rho_j, \forall i,j$](https://render.githubusercontent.com/render/math?math=%24%5Crho_i%3D%5Crho_j%2C%20%5Cforall%20i%2Cj%24)). The correspond Wald test is available in the `R` package `spsur`. 
-->

The SUR-SLM model can be estimated using maximum likelihood (@Lopez2014spatial) or instrumental variables (@Minguez2019).

Analysis
===================

@@

Figure \ref{fig:weekly-average-incidence-map} shows the geographical variation in the incidence of COVID-19 in Spain, as well as the temporal progression of the disease in weekly averages. Our analysis begins on March 13, which is the first date when every province had reported at least one case of COVID-19. It can be seen that the highest incidence at this early date was in the provice of Álava, in the North of Spain. While not the most populous province, with a population of only 331,549, Álava has the highest GDP per capita of all provinces. Vitoria, its main city, is the capital of the Basque Country and has been the focus of efforts, along with San Sebastian and Bilbao, to develop a "Global Basque City" [@Meijers2008strategic]. The other early focus of the pandemic in Spain was Madrid, which is the most populous province in the country and has the second highest GDP per capita after Álava. The early outbreaks in these two provinces can be traced throughout the progression of the pandemic over time, although by the end of the period under consideration, other provinces had matched and even surpased their incidence rates, including Segovia and Soria to the north of Madrid, and Ciudad Real and Albacete to the south.

@@

week11.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 11) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 13 - March 15") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week12.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 12) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds",
                       direction = 1) +
  ggtitle("March 16 - March 22") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", x = madrid_xy[1] - 9, y = madrid_xy[2] + 0.6) +
  geom_segment(x = madrid_xy[1] - 6, y = madrid_xy[2] + 0.5, xend = madrid_xy[1], yend = madrid_xy[2])

week13.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 13) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 23 - March 29") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid",
               x = madrid_xy[1] - 9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week14.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 14) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("March 30 - April 5") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

week15.plot <- covid19_spain %>%
  #filter(CCAA != "Canarias") %>%
  group_by(province, week = isoweek(Date)) %>% 
  summarise(ID_INE = first(ID_INE), mean_weekly_incidence = mean(Incidence)) %>%
  filter(week == 15) %>%
  left_join(provinces_spain,
            by = c("ID_INE")) %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = mean_weekly_incidence), color = "gray69") +
  scale_fill_distiller(name = "Mean Weekly Incidence", 
                       palette = "Reds", 
                       direction = 1) +
  ggtitle("April 6 - April 11") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom",
        plot.margin = margin(0, 0, 0, 0, "cm")) +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])

# Create grid layout
lay = rbind(c(1, 1, 2, 2),
        c(1, 1, 2, 2),
        c(3, 3, 4, 4),
        c(3, 3, 4, 4),
        c(NA, 5, 5, NA),
        c(NA, 5, 5, NA))

grid.arrange(week11.plot, week12.plot, week13.plot, week14.plot, week15.plot, layout_matrix = lay)
```

Figure \ref{fig:descriptives-temperature} shows the distribution of the environmental variables in Spain. For ease of visualization we aggregate the provinces by Autonomous Community. Each box-and-whisker in the figure represents the distribution of the variable for a community over the 30-day period. In the plot, the communities have been sorted by latitude, so that Principado de Asturias is the northernmost community, and Canarias the southernmost. As seen in the figure, there is a relatively wide range of values both within and between provinces over the 30-day period examined. The top panel of the figure shows the distribution of mean temperatures. The lowest mean temperature for a community on any given day was 2.8C, and the highest 22.4C, for a range of approximately 20 degrees. Likewise, there is a great deal of variability in humidity, as seen in the middle panel of the figure, where the lowest mean humidity for any community is 48.3 and the highest is 99.6. Finally, the bottom panel displays mean daily hours of sunshine in the community. This variable displays the most variability within communities over time, but remains relatively constant across communities. It is important to note that these are summaries by community, and the actual values of these variables for the provinces display somewhat more variability.

@@

# Summary of hours of sunshine by Autonomous Community
sunshine.summary <- covid19_spain %>% 
  group_by(CCAA, Date) %>%
  summarize(Value = mean(Sunshine_Hours)) %>%
  mutate(Variable = "Mean Daily Hours of Sunshina in Community")

# Bind tables
env.summary <- rbind(temp.summary, humidity.summary, sunshine.summary) %>%
  mutate(Variable = factor(Variable, 
                           levels = c("Mean Temperature in Community",
                                      "Mean Humidity in Community",
                                      "Mean Daily Hours of Sunshina in Community")))

# Plot
ggplot(data = env.summary, 
       aes(x = CCAA,
           y = Value)) +
  geom_boxplot() +
  theme_tufte() +
  theme(axis.text.x = element_text(angle = 90)) +
  xlab("Autonomous Community (sorted from north to south)") +
  ylab("Value") +
  facet_wrap(~ Variable, nrow = 3, 
             scales = "free_y")
```

Figure \ref{fig:map-controls} includes three maps that display the spatial variation of our control variables, namely GDP per capita, percentage of older adults in province, population density, and presence of mass transit systems. As seen there, GDP per capita is higher in Madrid and the northeast part of the country, mainly in Pais Vasco and Cataluña. Percentage of older adults is somewhat more checkered, with high values in Madrid and other provinces in the center-west part of the country, but also in some provinces in the north. Outside of provinces with major cities (e.g., Madrid; Bizkaia and Gipuzkoa in Pais Vasco; Pontevedra in Galicia), population density tends to be higher in provinces along the Mediterranean coast. The final panel in the figure shows the five provinces in the country that have higher order mass transit systems (e.g., metro).

```{r map-controls, cache=TRUE, echo=FALSE, fig.height= 9, fig.cap="\\label{fig:map-controls}Spatial distribution of control variables by province", warning=FALSE, cache = TRUE}
# GDP per capita:
q <- quantile(provinces_spain$GDPpc)
gdppc.plot <- provinces_spain %>%
  mutate(GDPpc = GDPpc,
         GDPpc = case_when(GDPpc < q[2] ~ "1st",
                             GDPpc > q[2] & GDPpc <= q[3]  ~ "2nd",
                             GDPpc > q[3] & GDPpc <= q[4] ~ "3rd",
                             GDPpc > q[4] ~ "4th")) %>%
  #filter(CCAA != "Canarias") %>%
  ggplot() +
  geom_sf(aes(fill = GDPpc), color = "gray69") +
  scale_fill_manual(name = "Quartile", values=c("#FFFEDE","#FFDFA2", "#FFA93F", "#D5610D")) +
  ggtitle("GDP per capita") +
  theme_bw(base_size = 9) +
  theme(axis.text = element_blank(),
        legend.position = "bottom") +
  geom_sf_text(label = "Madrid", 
               x = madrid_xy[1] - 9, 
               y = madrid_xy[2] + 0.6,
               family = "serif") +
  geom_segment(x = madrid_xy[1] - 6, 
               y = madrid_xy[2] + 0.5, 
               xend = madrid_xy[1], 
               yend = madrid_xy[2])
